%!TEX root = ../../template.tex

\section{Multi-Protocol Communication and Adaptive Protocol Selection}
\label{sec:multi_protocol}

Single-protocol (i.e. Wi-Fi only, \gls{BLE} only, \gls{LoRa} only, etc.) communication
architectures are susceptible to a fundamental vulnerability: if their chosen
medium becomes unavailable or inefficient due to interference, range
limitations, or infrastructure failures, the entire system loses connectivity
and most times ceases to operate altogether.

This becomes especially grave in disaster scenarios where communication
conditions change unpredictably due to factors such as \gls{RF} interference
caused by debris, obstacles blocking line-of-sight propagation, or
damaged/unavailable access points.

Existing approaches to multi-protocol \gls{iot} systems can be divided into
three categories: multi-channel resilience architectures that orchestrate
communication technologies for emergency scenarios; implementations that show
the feasibility of protocol heterogeneity on embedded platforms; and adaptive
selection frameworks that switch protocols based on runtime conditions.

\subsection{Multi-Channel Resilience Architectures}
\label{subsec:multi-channel}

The \gls{AWCT} framework~\cite{awct} orchestrates \gls{LPWAN} on top of
\gls{LoRaWAN} with ad-hoc networks (Bluetooth and Wi-Fi) specifically for
standby emergency communication. Their architecture adds three modules to
standard \gls{iot} devices (Raspberry Pi boards, in their test case): a battery
module for power management, a power interrupt handler that triggers emergency
mode when the power grid fails, and an ad-hoc bridge that forwards packets
between the Bluetooth, Wi-Fi and \gls{LPWAN} interfaces.

The system leverages dense \gls{iot} device deployment to provide emergency coverage,
demonstrating that existing infrastructure can still serve a purpose during
scenarios where the main power grid suffers issues.
However, AWCT's reliance on centralized \gls{LoRaWAN} gateways for Internet
connectivity creates a single point of failure when those gateways become
unreachable.\\

A more comprehensive heterogeneous approach is presented
in~\cite{heterogeneous-disaster}, which integrates \gls{HF} radio with
\gls{NVIS} technology, satellite links, \glspl{WSN}, and \gls{DTN} with mobile
drones for disaster monitoring. Their system uses \gls{RPL}~\cite{rfc6550} with
three separate instances to differentiate traffic by priority: human data
(voice/text via Bluetooth) receives the highest priority, followed by
drone-collected data and finally sensor data.

The \gls{NVIS} backhaul provides 250 km coverage radius without line-of-sight
requirements, offering a cost-effective alternative to satellite
communications.

While demonstrating successful real-world validation in Antarctica and urban
deployments, the architecture's core \gls{NVIS} topology with centralized coordination
contrasts fully distributed operational requirements. Furthermore, their \gls{WSN}
layer requires a minimum 20-second sending interval to maintain acceptable
packet loss rates at 10 hops, which highlights throughput limitations of
single-channel tree topologies.\\

Security-focused multi-channel approaches like MCSC-WoT~\cite{mcsc-wot} combine
\gls{AES} encryption with dynamic channel hopping across 2.4 GHz Wi-Fi channels
to defend against jamming and eavesdropping attacks. Their lightweight
synchronization mechanism minimizes energy consumption while maintaining
security through \gls{FHSS} patterns generated via \gls{PRNG}. Nodes that lose
synchronization can rejoin by hopping to random channels and waiting for the
next synchronization signal.

However, the system depends on a master node broadcasting synchronization
signals and uses an initial \gls{PRNG} seed shared among all nodes, raising
questions about scalability and seed distribution mechanisms -- particularly
how new nodes can acquire seeds if deployed to a network some time after the
system is operational.

\subsection{Protocol Heterogeneity on Embedded Platforms}
\label{subsec:proto-het}

The work presented in~\cite{multiproto_gateway} shows the technical feasibility
of multi-protocol operation on commodity Wi-Fi/\gls{BLE} modules by integrating
\gls{ESP-NOW}, \gls{ZigBee}, and \gls{Modbus} protocols on devices from the
ESP32 family to construct multi-hop, tree-based wireless networks.

In this tree-based architecture, devices organize hierarchically with a central
gateway serving as the root. Their implementation uses \gls{BLE} advertising
beacons for neighbor discovery with \gls{RSSI} measurements (for distance
estimation), computing parent selection priorities based on weighted
combinations of child count, \gls{RSSI} values, and hop count in the network
tree of a given node. The system also supports automatic parent reselection
when the current parent becomes unreachable, making it somewhat resilient to
single node failures.

Testing demonstrated successful multi-hop operation up to 5 hops in office
environments, but evaluation was limited to linear network topologies.
Additionally, this architecture relies on a tree topology with parent/child
communication that is centered around a gateway (coordinator) rather than being
a \gls{P2P} mesh, creating a single point of failure at the coordinator node.\\

\subsection{Runtime Adaptive Protocol Selection}
\label{subsec:runtime-adaptive}

%%%%%%%%%%%%%%%%%%%%%%%%%
% A recent approach to protocol adaptation is presented in~\cite{adaptive2025},
% which introduces a closed-loop control system with four integrated components:
% a context monitor sampling runtime metrics at 1 Hz, a decision engine using
% multi-criteria weighted scoring across six dimensions (message frequency,
% payload size, network conditions, packet loss rate, energy budget, \gls{QoS}
% requirements), protocol adapters encapsulating protocol-specific libraries, and
% a learning component that adjusts thresholds using \gls{EWMA} for
% deployment-specific patterns.
%
% Some key aspects of this work include \gls{hysteresis} control with a threshold
% band to prevent oscillation between protocols, and switching cost awareness
% requiring benefits to exceed costs by a certain threshold before transitioning.
%
% However, this framework assumes infrastructure availability for protocol
% endpoints (\gls{MQTT} brokers, \gls{HTTP} servers, \gls{CoAP} endpoints) and was
% evaluated only on laptop platforms using Python libraries rather than embedded
% C implementations.
%
% Nevertheless, the evaluation methodology provides valuable insight into the
% energy expenditure of these different approaches (further explored in
% Section~\ref{subsec:receiver_energy}), and grants useful foundations for more
% complex multi-protocol approaches, in particular with the multi-criteria
% decision framework and \gls{hysteresis} control mechanism, but the actual
% system cannot operate when infrastructure fails.\\
% %%%%%%%%%%%%%%%%%%%%%%%%%

The MINOS platform~\cite{minos2019} exemplifies the limitations of centralized
multi-protocol approaches. While providing sophisticated multi-protocol support
(CORAL-SDN and Adaptable-RPL) with dynamic protocol deployment and real-time
parameter tuning, the system depends fundamentally on a centralized \gls{SDN}
controller, \gls{MQTT} broker, and web server infrastructure. The architecture's
single point of failure means that when the controller becomes unreachable the
entire system loses its adaptive capabilities and reverts to static operation
at best, or complete failure at worst.
\todo{FIXED: removed MDPI paper}

\subsection{Discussion}
\label{subsec:rw_discussion_3}

Existing frameworks demonstrate the technical feasibility of heterogeneous
channel/protocol coordination but rely on centralized control mechanisms
incompatible with disaster scenarios. AWCT~\cite{awct} depends on centralized
\gls{LoRaWAN} gateways for Internet connectivity, the heterogeneous disaster
architecture~\cite{heterogeneous-disaster} uses centralized \gls{NVIS} coordination
with minimum 20-second send intervals at 10 hops, MCSC-WoT~\cite{mcsc-wot}
requires master nodes broadcasting synchronization signals, the multi-protocol
gateway~\cite{multiproto_gateway} uses tree topologies with gateway
coordinators as single points of failure, and MINOS~\cite{minos2019} assumes
SDN controllers with persistent \gls{MQTT}/web infrastructure.\\

% While adaptive protocol selection frameworks~\cite{adaptive2025} provide
% runtime switching strategies, they assume infrastructure availability
% (\gls{MQTT} brokers, \gls{HTTP} servers, \gls{CoAP} endpoints) and have not
% been validated on embedded platforms.\\

We have found no work that addresses fully distributed multi-protocol
coordination where nodes can autonomously select and switch protocols without
resorting to persistent infrastructure.

\todo{REMOVED: ubabel's approach will...}
% \gls{ubabel}'s approach to decentralized protocol selection is detailed in
% Section \ref{component:protocol_manager} along with masterless synchronization
% mechanisms.
