%!TEX root = ../../template.tex

\section{Operating Conditions and Failure Modes}
\label{sec:operating_conditions}

\gls{ubabel} is designed to remain operational under conditions where
traditional infrastructure becomes unreliable or unavailable. This section
establishes the operational environment and failure scenarios that drive our
architectural decisions.

\subsection{Target Operating Environments}

\gls{ubabel} targets deployments where infrastructure availability cannot be
guaranteed, and in which autonomous operation is a necessity:

\paragraph{Disaster Scenarios:} Locations susceptible to natural disasters
(i.e., earthquakes, floods) or infrastructure failures that eliminate
power grids and network access points.

\paragraph{Remote or Hazardous Locations:} Industrial sites, wilderness
monitoring, or other dangerous environments where manual intervention and
infrastructure maintenance are impractical or impossible.

\paragraph{Local-First Deployments:} Environments where data sovereignty
requirements or privacy concerns require local processing and control rather
than cloud dependency, even when connectivity is available.
\todo{add more focus on smart home stuff?}

\subsection{Expected Failure Modes}

The architecture must maintain functionality under the following conditions:

\paragraph{Infrastructure Collapse:} Complete loss of access points and
external network connectivity. Devices must self-organize and maintain
\gls{P2P} communication without infrastructure support.

\paragraph{Communication Medium Disruption:} Individual radio interfaces may
become unavailable due to interference, jamming, congestion, or physical
obstacles blocking signal propagation. Technology diversity provides
alternative communication paths and ensures continued operation.

\paragraph{Network Partitioning:} Physical damage or communication range
limitations may split networks into isolated subgroups. Devices within
partitions must continue autonomous operation while supporting eventual
reconciliation when connectivity is restored.

\paragraph{Node Failures:} Individual devices may become unavailable due to
battery depletion, physical damage, or environmental conditions. The network
must maintain operation despite individual node failures without depending on
specific coordinator nodes.

\paragraph{Intermittent Connectivity:} Communication links may be unstable with
varying packet loss, latency, and availability. Protocols must tolerate
unreliable links and adapt to changing conditions.
